(WT1) Tree establishment in models of forest dynamics: a key priority for further research  \n(WT2) Tree establishment processes in models of forest dynamics: capturing the drivers remains challenging

### 2022-12-14 manuscript/manuscriptRW.Rmd
changes based on feedback and discussion with Harald on 12-12-2022

---
title: Drivers of tree regeneration remain challenging to be captured by models of
  forest dynamics
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Mats Mahnken,
- Samuel Fischer,
- Xavier Morin,
- Manfred Lexer,
- Christian Hochauer,
- Katarina Merganicova,
- Tobias Mette,
- Werner Rammer,
- Kristin H. Braziunas,
- Paola Mairota,
- Josef Bruna,
- Heike Lischke,
- Daniel Scherrer,
- Simon Scheiter,
- Thomas Hickler,
- Tim Anders,
- Jessica Hetzer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
  bookdown::word_document2: default
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: draft.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree regeneration is a key process in forest dynamics, particularly in
the context of forest resilience after disturbance and climate change.
Models of forest dynamics have been in use since more than 50 years, but
an assessment of their capacity to properly represent regeneration is
still lacking. Using 15 models and model variants that were build to
capture long term (decadal to centennial) forest dynamics at the stand,
landscape and global levels, we simulate tree regeneration across at 200
sites representing environmental gradients across Europe. The simulation
results are benchmarked against data from forest reserves collected at
these 200 sites. We aim to assess how well the models of forest dynamics
are capturing the overall abundance, species composition and mortality
in tree recruitment. The results show that most of the models tend to
over-estimate recruitment rates although this is not always connected to
large mortality rates in the early stages of population dynamics. In the
models, the effect of light availability on recruitment rates is
captured better than the effect of temperature and soil moisture, but
the patterns are not consistent across models. Species richness matches
quite well compared to the ranges observed in the data. Models
simulating higher species diversity at the stand level do not feature
higher diversity in recruitment. *Abies alba* was well captured across
temperature and soil moisture gradients by most models, whereas *Fagus
sylvatica* was generally underestimated. *Picea abies* recruitment was
generally over-estimated, particularly at colder sites. HERE TWO
SENCENTECES DISCUSSION/CONCLUSION. HERE A SO WHAT SENTENCE

@@

Forests provide a wide range of products and services of vital
importance. Under the pressure of climate change and increasing
disturbances, it is becoming ever more important to understand how
forests structure, composition and functions will change, and evaluate
forest capacity to adapt or be resilient after disturbance
[@lindner2010]. A wide range of models of forest dynamics were developed
over the past decades considering the impacts of climate on forest
dynamics [@bugmann2022]. From these studies, it is becoming increasingly
evident that while we have a reasonably good understanding how to model
tree growth [@vanclay1997; @bugmann1996], and substantial efforts have
been dedicated to improving the depiction of tree mortality processes
[@bugmann2019], tree regeneration remains a much less studied process
that is often represented rather coarsely in these models, which is a
clear research gap in the context of climate-induced forest disturbances
and forest resilience.

Tree regeneration depends on multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001]. All these processes are all difficult
to assess, and some of them are scarcely understood and thus appear
highly stochastic [@bogdziewicz2021]. Identifying the appropiate level
of complexity for the formulation of the key factors that are leading to
successful tree recruitment is challenging. By tree recruitment, we
refer to the passing of trees to selected diameter threshold (ingrowth)
as a definition of successful regeneration [@käber2022]. Overall, models
are needed to (1) synthesize existing empirical data and explore their
consequences, and (2) assess future tree recruitment e.g. in the context
of global change scenarios. Currently, regeneration processes in dynamic
forest models are handled in a multitude of ways [@konig_tree_2022;
@bugmann2022], from entirely ignoring tree regeneration (as done in
classical forest growth models, e.g. @pretzsch2002), across the use of a
few simple environmental filters, as done in most forest "gap" models
(e.g., @shuman2011) and Dynamic Global Vegetation Models (e.g.,
@smith2001; @hickler2012), to complex approaches that incorporate local
feedback from the canopy, multiple ecological processes and often also
small time steps (e.g., @wehrli2006; @seidl2012), or field-based
statistical parameterizations, which however are not easy to extrapolate
in space and time (e.g., @ribbens1994).

Given the current strategies that are used in models of forest dynamics
to represent tree recruitment, model behavior is facing several problems
and that can have strong impacts on forest dynamic, such as very high
levels of recruitment, which then necessitate excess mortality to early
stages to arrive at correct stand structure and composition. Also, often
inadequately high species diversity in regeneration is simulated. This
is characteristic of \'classical\' forest gap models [@gutiérrez2016],
at least as long as the simulations comprise a multitude of species.
Other models use calibration against local canopy-level data to define
appropriate recruitment levels, which is likely to strongly hamper the
general applicability of these models e.g. under scenarios of climate
change.

Correctly capturing the species composition in recruitment as a function
of the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[@lischke2006]. A related issues is the excessive reduction of species
richness due to positive feedback effects such that eventually just
single-species forest stands remain. This is sometimes handled by the
incorporation of a low level of seed influx of all species at all times
(e.g., @lischke2006 ). However, simulated species composition is usually
exceedingly sensitive to that level, whereas the parameters of such
functions are poorly constrained by field data. Furthermore, there is
also often a problem with the insufficient data-based constraint on
parameter values for models that start from very small tree sizes (e.g.,
10 cm of height or even seeds) and track tree development in a
process-oriented manner by considering a multitude of ecological
influences, rather than being statistically driven and emphasizing
recruitment into a larger size class (e.g., @wehrli2006 ).

Thus, a focus on the modeling of tree regeneration processes is sorely
needed and long overdue (cf. @price2001) if we are to make reliable
projections of future forest dynamics, i.e. when the models need to be
operated in extrapolation mode, but also from a fundamental ecological
point of view (increasing systems understanding).

Also in the field of ecology, tree recruitment has not received a lot of
attention [@hanburybrown_forest_2022], although many sub-processes have
been studied in detail. Yet, little data is available that covers all
regeneration processes within one species along environmental gradients,
let alone for a vast suite of species. Thus, the available data are
fragmented, which constitutes a major problem for model building. On the
one hand, monitoring on permanent sample plots (such as National Forest
Inventories) often measure tree recruitment [@zell_modeling_2019],
however often with a design that captures the regeneration intensity
after the stand initiation phase in productive forests [@hallsby2015].
Data from permanent plots, although available across regions, are also
highly heterogeneous, rendering their utility difficult in a model
context (cf. @käber2021). On the other hand, targeted studies to measure
tree recruitment on experimental sites are often limited in spatial
extent, species studied, or sub-set of regeneration processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale multi-species harmonized data set
on tree recruitment [@käber2022]. We included models that are based on a
range of 'philosophies', including models operating at the stand to the
global scale as well as the range of models from empirically derived to
"process-based" (e.g. @fabrika2012; @mette2014; @lexer2001;
@bugmann1996). Because of the large variability in recruitment patterns
in nature and the large amount of factors driving recruitment, including
some that are not included in most models, such as deer browsing, we do
not aim at a detailed statistical evaluation of each model. Instead, we
aim at evaluating the general recruitment patterns and magnitudes
simulated by the models, and benchmark the simulated regeneration niche
of multiple species against empirical data along a wide environmental
gradient (temperature, moisture, and light availability).

More specifically, we aimed to answer the following questions. 1) Are
models of forests dynamics capturing correctly tree recruitment levels
across environmental gradients? 2) Are models of forest dynamics
representing correctly the initial composition of tree populations? And
3), do the models of forest dynamics correctly capture the mortality in
tree in-growth?.

@@

Fifteen models of forest dynamics (including two model featuring
variants) were used to assess tree regeneration. The selected models
simulate forest dynamics at the stand, landscape and global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ, with most of the models featuring a largely
"process-based" approach, whereas two models based on formulations
derived from the statistical analysis of inventory data (SYBILA and
xComp). Some of the models largely rely on the original approach
underlying forest gap models (ForClim 1, ForClim 11, LandClim, Landis-II
); whereas some include plant ecophysiological processes (FORMIND,
PICUS, 4C, iLand). These different approaches underlying the models also
has strongly influenced the formulation of tree recruitment.

Based on their establishment modules, the models can be categorized as
regeneration or recruitment (or ingrowth) models [@vanclay1994;
@konig_tree_2022] (see Table \@ref(tab:tableModels)). Regeneration
models include processes such as flowering and pollination, seed
production, seed dispersal, germination or seedling establishment that
lead to the simulated number of seedlings or saplings. Recruitment
models, in contrast, introduce a certain number of seedlings or saplings
with certain characteristics such as biomass or diameter, without
considering earlier development processes. We can further distinguish
models that feature a feedback in the regeneration, i.e. where the level
and species composition of regeneration is influenced by the existing
adult tree community via the production of seeds, seedlings or saplings.

The starting point for regeneration in the models also differs, ranging
from seedbank, seed, seedling to sapling (i.e., trees much larger than
10 cm height, often at least 2-3 m tall), which implies that the larger
size threshold, the larger the number of ecological processes that must
be aggregated via parameterizations because they cannot be treated
explicitly. This reduces model complexity but comes at the cost of
blurring process representation.

The overall complexity in the establishment modules also varies
considerably. Following @bugmann2022, we can classified seven models
(iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS, ForCLim and TreeMig) as
having rather high complexity in their tree recruitment modules (mean
complexity across all the recruitment process with value above zero in
Table \@ref(tab:tableModels)). The highest complexity is when defining
processes such as the process used for modeling establishment or how is
frost occurrence taken into account in establishment,

Two models, ForClim and ForCEEPs used two alternative model versions.
ForClim variant 1 [@bugmann1996] is based on an establishment module
that adheres closely to the concept introduced by @botkin1972, whereas
ForClim variant 11 is adopting a rather different approach for
establishment model that considers
space filling along with different \'points of entry\' for the action of
specific
environmental factors @huber2020. These two model variants allow us to
evaluate a more process-based establishment module (variant 11) against
a simpler module (variant 1) while the rest of the model structure
remains the same. Similarly,
the two ForCEEPS variants used here allow us to isolate the importance
of the
canopy feedback via seed trees for the quantity and quality of simulated
regeneration, as one variants includes this feedback whereas the other
one does
not.

@@

kable(modelDF, 
      #format = "markdown",
      format = "latex", booktabs = TRUE,
      caption = "List of models included.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "5em") |> 
      kableExtra::column_spec(column = 3, width = "5em") |> 
      kableExtra::column_spec(column = 4, width = "5em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "5em") 
```

@@

Recruitment data covering a wide range of environmental conditions are
hard to obtain, and this is one of the reasons why most of the models of
forest dynamics have never been confronted with a data-set covering such
gradients over a large number sites, to evaluate how well they are
capturing recruitment. The recruitment observations used here are from a
novel and unprecedented network of sites in forest reserves that
represent the range of environmental gradients in temperature and
precipitation in Europe compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). All these forests have not been
managed for at least 10 years prior to their designation as reserves,
and most of them provide time series of natural forest dynamics over
multiple decades and up to ca. 80 years. The censuses periods range from
37 to 155 years, with most of sites with an average of 14 years. The
plot sized ranged from 0.02 ha to 5.52 ha, the recruited values were
further processed and aggregated to provide recruited values per ha
following @käber2022. The data provide information at the tree level,
thus providing information for sequential comparisons such as
recruitment and death between measurements.

We selected 200 sites from this network as the benchmarking data set for
the simulation exercise. Each of these sites features at least two
consecutive measurements. Regeneration was measured at threshold of 7 or
10 cm of stem diameter at all the sites, except for 35 sites where only
regeneration at the 10 cm diameter threshold was available. The
observations used in this study featured 30.900 recruited trees, and
recruitment values per site and sample ranging from zero to 1246 trees
per ha and a mean of 56 trees per ha recruited. For more details on this
unique dataset and the detailed recruitment information please refer to
@käber2022.

@@

The overarching goal of this experiment was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. To this end, each moding group
was provided with a detailed protocol (available in the Supplementary
Material X, henceforth "SM") with instructions how to perform the
simulations, providing input variables on climate and soil conditions,
the specification of the simulation setup and the expected output
variables. Neither site information (except for the data specified
below) nor any data on tree regeneration were provided prior to the
simulation. That is, the models were run in "blind flight" mode.

The input variables were collected and aggregated to be adapted to the
different model demands. The time series of climatic variables from 1981
to 2018 were provided in hourly (Era5-land data, @muñozsabater2019),
daily and monthly resolution (CHELSA data, @karger2021). Some variables
required for some models such as relative humidity or vapor pressure
deficit had to be calculated from these variables. The final strategy on
how to use the climate data was different in each model based on the
approach that best suited the model (Table \@ref(tab:tableModels)). Soil
quality data was provided as a continuous value between 1 and 5
(Soilgrids dataset, @hengl2017, @poggio2021). The protocol also provided
the elevation, slope and aspect for each of the 200 sites
[@asterscienceteam2019], no other spatial information such as
coordinates were provided with the exception of iLand which used the
blurred coordinates to derive highly detailed soil information values
for the 200 sites.

The simulations were run in the absence of natural disturbances, only
the model LPJ-GUESS had to include HERE LPJ-GUESS PROVIDE [ONE
SENTENCE]{.underline} OF WHY YOU INCLUDED DISTURBANCES AND WHAT TYPE.
The simulations were set up to sample species-specific regeneration
rates per decade and per ha in the equilibrium state of the model,
typically entailing a \"spin-up\" run (as we did not provide any forest
data). The modeling teams decided on the simulated area and how they
derived these samples. The exact length of the simulation was also
decided by the modeling teams (Table \@ref(tab:tableModels)). Further
details on how each modeling team prepared the simulations and the
outputs are available in the SM 2.

The simulations were run in the absence of management to a
simulated equilibrium ("Potential Natural Vegetation") with the current
climate, as the ultimate goal was to derive the regeneration niche. This
entails the assumption that (1) the observations from the forest
reserves reflect no traces of forest management and (2) an equilibrium
between forest dynamics and climate. While the former is likely to be
met at least in many of the EuFoRIA reserves, the latter may be more
debatable. However, in the absence of detailed data on the history of
each plot in the EuFoRIa network, some assumptions had to be made. We
define the regeneration niche as the passing of a diameter threshold of
7 or 10 cm respectively (synonym: ingrowth). Both the width of the
regeneration niche (i.e., in environmental space) as well as the
intensity of the recruitment process (i.e., the number of ingrowth trees
per area and per unit of time) were of interest.

>>>>

---
title: "(WT1) Tree establishment in models of forest dynamics: a key priority for further research  \n(WT2) Tree establishment processes in models of forest dynamics: capturing the drivers remains challenging"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience after disturbance and climate change.
Models of forest dynamics have been in use since more than 50 years, but
an assessment of their capacity to represent tree establishment is
lacking. We assess how well current models are capturing the overall
abundance, species composition and mortality of tree establishment.

Using 15 models and model variants buil to capture longterm (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are benchmarked
against comprehensive data from forest reserves.

Most of the models overestimate ingrowth rates, but this is not always
compensated by high simulated mortality rates in the early stages of
population dynamics. In the models, the effect of light availability on
ingrowth rates is captured better than the effect of temperature and
soil moisture, but the patterns are not consistent across models.
Simulated species diversity matches the observed ranges quite well.
Models simulating higher species diversity at the stand level do not
feature higher ingrowth diversity. The ingrowth of *Abies alba* is
captured across temperature and soil moisture gradients, whereas *Fagus
sylvatica* is generally underestimated. *Picea abies* ingrowth is
generally overestimated, particularly at colder sites. Increasing
complexity in the tree establishment modules of the models is not
related to higher accuracy of simulated tree establishment. Furthermore,
individual model design is more important than scale (stand, landscape,
global) and approach (empirical, process-based) to accurately capture
tree establishment.

Even though the overestimation of tree establishment represent a
challenge
for the usefulness of model~~s~~ projections under uncertain conditions
such a climate change, and although some formulations should be
reconsidered to correct excessive ingrowth levels, it is remarkable that
most of the~~se~~ models capture the essential features of the highly
complex process of tree establishment, while not having been
parameterized with such a data. We conclude that in spite of all the
uncertainties, many models hold promise for assessing longterm forest
dynamics under changing environmental conditions.

@@

Forests provide a wide range of products and services of vital
importance to humankind.
Under the pressure of climate change and increasing disturbances, it is
becoming ever more important to understand how forest structure,
composition and function will change, and to evaluate forest capacity to
adapt or be resilient after disturbance [@lindner2010]. A wide range of
models of forest dynamics were developed over the past decades
considering the impacts of climate on forest dynamics [@bugmann2022].
From these studies, it is evident that while we have a reasonably good
understanding how to model tree growth [@vanclay1997; @bugmann1996], and
substantial efforts have been dedicated to improving the depiction of
tree mortality [@bugmann2019], tree establishment remains a much less
studied process that is often represented rather coarsely in these
models, which is a clear research gap in the context of climate-induced
forest disturbances and forest resilience.

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001]. All these processes are difficult to
assess, and some of them
are scarcely understood and thus appear highly stochastic
[@bogdziewicz2021]. Identifying the appropriate level of complexity for
the formulation of the key factors that are leading to successful tree
establishment is challenging. Currently, tree
establishment processes in dynamic forest models are handled in a
multitude of
ways [@konig_tree_2022; @bugmann2022], from entirely ignoring it [as
done in classical forest growth models, e.g., Pretzsch et al.
-@pretzsch2002], across the use of a few simple environmental filters,
as done in most forest \"gap\" models [@shuman2011] and Dynamic Global
Vegetation Models [e.g., @hickler2012; @smith2001], to complex
approaches that incorporate local feedback from the canopy, multiple
ecological processes and often also small time steps [e.g., @seidl2012;
@wehrli2006], or field-based statistical parameterizations, which
however are not easy to extrapolate in
space and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest
dynamics to represent tree establishment, model behavior is often facing
problems that can have strong impacts on simulated forest dynamics, such
as very high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to arrive at correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
\"classical\" forest gap models [@gutiérrez2016], at least as long as
the simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to
unrealistic successional drift in the model, which must be corrected by
factors that are hard to parameterize [e.g., @lischke2006]. A related
issue is the excessive reduction of species richness due to positive
feedback effects such that eventually just single-species ~~forest~~
stands remain. This is sometimes corrected by the incorporation of a low
level of seed influx of all species at all times [@lischke2006].
However, simulated species composition is usually exceedingly sensitive
to that level, whereas the parameters of such functions are poorly
constrained by field data. Lastly, there is often a problem with
insufficient observational constraint on parameter values for models
that start from very small tree sizes (e.g., 10 cm of height or even
seeds) and track tree development in a process-oriented manner by
considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001] if we are to make reliable projections
of future forest dynamics, i.e., when the models need to be operated in
extrapolation mode, but also from a fundamental ecological point of view
(increasing systems understanding). Also in the field of general
ecology, tree establishment has not received a lot of attention
[@hanburybrown_forest_2022], although many sub-processes have been
studied in detail. Yet, little data is available that covers all the
processes within one species along environmental gradients, let alone
for a vast suite of species. Thus, the available data are fragmented,
which constitutes a major problem for model building. On the one hand,
monitoring on permanent sample plots (such as National Forest
Inventories) often measure tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation
phase[@hallsby2015]. Data from permanent plots, although available
across regions, are highly heterogeneous, rendering their utility
difficult in a modeling context[@käber2021]. On the other hand, targeted
studies to measure tree establishment on experimental sites are often
limited in spatial extent, species studied, or the sub-set of processes
that are targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold (\"ingrowth\")
and thus as successful tree establishment. We included models that are
based on a range of \"philosophies\", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to \"process-based\" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not included in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.

More specifically, we aimed to answer the following questions. 1) Are
models of forest dynamics capturing accurately tree ingrowth levels
across environmental gradients? 2) Are they matching the initial species
composition of the tree ingrowth layer? And 3), do they correctly
represent the mortality in tree establishment?.

@@

Fifteen models of forest dynamics (including two model featuring
variants) were used to assess tree establishment. They simulate forest
dynamics at the stand, landscape and global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely \"process-based\" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SYBILA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and
LandClim)[,]{.underline} whereas some [a]{.underline}re based on plant
ecophysiological processes (e.g., FORMIND and iLand). The different
approaches underlying the models have strongly influenced the
formulation of tree establishment.

The establishment modules can be categorized as \'regeneration\' or
\'recruitment\'
models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination
seedling [g]{.underline}rowth, which ultimately lead to the simulated
number of trees. Recruitment models, in contrast, introduce a certain
number of new trees with certain characteristics such as biomass or
diameter, without explicitly considering earlier development processes.
We can further distinguish models that feature a feedback in the
establishment, i.e., where the level and species composition
of establishment is influenced by the existing adult tree community via
the
production of seeds, seedlings or saplings, from models that do not
contain such feedback.

The starting point for in the models differs as well, ranging from
seedbank, seed or seedling to sapling (i.e., trees much larger than 10
cm height, often at least 2-3 m tall).This implies that the larger the
size threshold, the larger the number of ecological processes that must
be aggregated via parameterizations because they cannot be treated
explicitly. This reduces model complexity but comes at the cost of
blurring process representation.

The overall complexity in the establishment modules also varies
considerably. Following @bugmann2022, we can classify seven models
(iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS, ForClim and TreeMig) as
having rather high complexity in their establishment modules (average
complexity across all processes \>0, Table \@ref(tab:tableModels)). In
most models[,]{.underline} the two processes in the establishment
modules with the highest complexity are the statistical approach used to
define the successful number of new treesand how frost occurrence is
taken into account.

Two models, ForClim and ForCEEPs, were used here with two alternative
variants.
ForClim variant 1 [@bugmann1996] is based on a recruitment module that
adheres closely to the concept introduced by @botkin1972, whereas
ForClim variant 11 is adopting a rather different approach that
considers space filling along with different \'points of entry\' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback via seed trees for the
quantity and quality of simulated ingrowth[,]{.underline} as one variant
includes this feedback whereas the other does not.

@@

#there is a better way of doing footnotes, have a look here https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf and change
names(modelDF)[2] <- paste0(names(modelDF)[2], "(a)")
names(modelDF)[3] <- paste0(names(modelDF)[3], "(b)")
names(modelDF)[4] <- paste0(names(modelDF)[4], "(c)")
names(modelDF)[5] <- paste0(names(modelDF)[5], "(d)")
names(modelDF)[6] <- paste0(names(modelDF)[6],"(e)")
names(modelDF)[11] <- paste0(names(modelDF)[11],"(f)")

kable(modelDF, 
      #format = "markdown",
      format = "latex", booktabs = TRUE,
      caption = "List of models included.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "3em") |> 
      kableExtra::column_spec(column = 3, width = "3em") |> 
      kableExtra::column_spec(column = 4, width = "4em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "5em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical", 
                                 "C-Cohort, I-Individual",
                                 "RG-Regeneration, RC-Recruitment",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

                 notation = "alphabet") 
```

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number sites, to evaluate how well ingrowth is
captured. The observations used here are from a novel and unprecedented
network of sites in forest reserves that represent the range of
environmental gradients in temperature and precipitation in central
Europe as compiled in the framework of the EuFoRIa network[@euforia2019]
(Figure \@ref(fig:map)). All these forests have not been managed for at
least 10 years prior to their designation as reserves, and most of them
provide time series of natural forest dynamics over multiple decades and
up to ca. 80 years. The census periods range from 3 to 37 years, with an
average of 14 years. The data provide information at the tree level,
thus allowing for sequential comparison such as individual-tree ingrowth
and death between the measurements.

We selected 200 sites from this network as the benchmarking data set for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
kmeans clustering to define 200 clusters of plots from the
original set of 869 plots along the environmental dimensions of
temperature, water balance, soil quality and exposition (i.e., slope and
aspect). Each of these sites
featured at least two consecutive measurements. Ingrowth was measured at
threshold of 7 or 10 cm of stem diameter at all sites except for 35
sites where only ingrowth at the 10 cm diameter threshold was available.
Plot size ranged from 0.02 ha to 5.52 ha, and the data were further
processed and aggregated following @käber2022 to provide ingrowth rates
per ha and per decade. The observations used in this study featured
30.900 established trees, and ingrowth rates per site, sample and
decade ranged from zero to 1246 trees per ha, with a mean of 56 trees
per ha. For more details on this unique dataset and the detailed
establishment information, cf. @käber2022.

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a diameter threshold of 7 or 10
cm[,]{.underline} respectively (synonym: ingrowth). To this end, each
modeling group was provided with a detailed protocol (cf. Supplementary
Material 1, henceforth \"SM\") with instructions how to perform the
simulations, providing input variables on climate and soil conditions,
and the expected output variables. Neither were further site information
(except for the data specified below) nor any data on tree ingrowth
provided prior to the simulation. That is, the models were run in
\"blind flight\" mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models such as
relative humidity or vapor pressure deficit had to be calculated from
these variables. The final strategy how to use the climate data was
different in each model, based on the approach that best suited the
model (Table \@ref(tab:tableModels)). Soil quality data was provided as
a continuous value between 1 and 5 [Soilgrids dataset, Hengl et al.
-@hengl2017, Poggio et al. -@poggio2021]. The protocol also provided the
elevation, slope and aspect for each of the 200 sites
[@asterscienceteam2019], but no other spatial information such as
coordinates was provided, with the exception of iLand[,]{.underline}
which used the blurred coordinates to derive highly detailed soil data.

The simulations were run in the absence of natural disturbances. Only
the model LPJ-GUESS had to include HERE LPJ-GUESS PROVIDE [ONE
SENTENCE]{.underline} OF WHY YOU INCLUDED DISTURBANCES AND WHAT TYPE.
The simulations were set up to sample species-specific ingrowth rates
per decade and per ha in the equilibrium state of the model, typically
entailing a "spin-up" run (as we did not provide any forest data). The
modeling teams decided on the simulated area and how they derived these
samples. The exact length of the simulation was also decided by the
modeling teams (Table \@ref(tab:tableModels)). Further details on how
each modeling team prepared the simulations and the outputs are
available in SM 2.

The simulations were run in the absence of management to a
simulated equilibrium ("Potential Natural Vegetation") with the current
climate, as the ultimate goal was to derive the regeneration niche. This
entails the assumption that (1) the observations from the forest
reserves reflect no traces of forest management and (2) there is an
equilibrium between forest dynamics and climate. While the former is
likely to be met at least in many of the EuFoRIA reserves, the latter
may be more debatable. However, in the absence of detailed data on the
history of each plot in the EuFoRIa network, some broad assumptions had
to be made. Both the width of the regeneration niche (i.e., in
environmental space) as well as the intensity of the ingrowth process
(i.e., the number of ingrowth trees per area and per unit of time) were
of interest.

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* Two models included
their standard set of species for the simulations which is much larger
(i.e., ForClim 1, ForClim 11). This also applies to TreeMig, but in this
model *Betula pendula* was not included because HERE **TREEMIG**
PROVIDES THE END OF THIS SENTENCE SAYING WHY B pendula WAS NOT INCLUDED.
In three models, fewer than these elveven species were simulated (4C,
xComp, LPJ-GUESS) (Table \@ref(tab:tableModels)). HERE **4C** PROVIDES
[ONE SENTENCE]{.underline} WHY THEY SIMULATED LESS SPECIES. HERE
**XCOMP** PROVIDES [ONE SENTENCE]{.underline} WHY THEY SIMULATED LESS
SPECIES. HERE **LPJ-GUESS** PROVIDES [ONE SENTENCE]{.underline} WHY THEY
SIMULATED LESS SPECIES. Finally, the model aDGVM2 is based on a suite of
broad Plant Functional Types rather than individual species, and hence
this model used its default approach of PFTs for the simulations.

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1-ha plot in the equilibrium over time, or (2)
sampling several 1-ha plots from the simulated forest at one specific
point in time (in the equilibrium). This resulted in 880.000
observations per model that simulated the 11 species included in the
protocol (200 sites, 200 samples per site, 11 species and two diameter
thresholds). For the models that simulated additional species, their
ingrowth rates were aggregated as \"others\".
